Historic collections as a tool for assessing the global pollination crisis

### 2018-04-06 submitted/SupMat3.Rmd
Last touches

---
title: Historic collections as a tool for assessing the global pollination crisis
author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
csl: data/philb_style.xml
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
bibliography: data/references.xml
---

@@

For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1). Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., Bombus dalbhomi; @Morales2013).

@@

Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size. This correction accounts for the species-area relationship. Warmer colours indicate higher bee diversity. Note that some African countries may have incomplete listed faunas and that Alaska is included with US values. Countries with available historical changes in (b) bee, (c) syrphid and (d) butterfly richness within the last 100 years. Warmer colours indicate steeper average declines. Countries without data are coloured in white.

@@

#average richness change
colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- (heat.colors(10))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

Figure 2. Exploration of available data for bee records showing a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country. The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red. For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is similar in both time periods (b). A more careful analysis of this data would help complete the map of global declines (c). Note that this data are likely to contain strong undetected biases, as we explore below.

@@

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv("data/spain_clean_used.csv")
#Repeat richness and genus
rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#rich_tot #wow.
#extract genus
pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
dat$genus <- substr(dat$matched_name2, 
                    start = 1,
                    stop = pos-1)
genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot #wow, lot's of genus lost!
#genus_tot2
#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot3
#colSums(genus_tot3[,2:3])
#write.csv(genus_tot3, "raw_data/spain_species_per_genus.csv")

@@

Figure 3. Comparision of historic collections (1980's) and modern re-surveys (2016) of the rarefied richness of bees at six Spanish localities.

@@

Figure 4. Changes in rarefied species richness for different pollinator groups in New Zealand over time. All trends were non-significant (α = 0.05).

@@

Figure 5. Model estimated changes (± 1 SE) in the relative occurrence frequency of different New Zealand bee and fly species in museum collections over time.

@@

We thank Curro Molina, Carola Warner, Patrick McQuinn, and Crona McMonagle for data entry and Gregorio Aguado for carrying out the Spanish re-sampling. We thank Barry Donovan for providing New Zealand bee collection records and E. Asensio for sharing his historical data and knowledge. We thank the "Museo Nacional de Ciencias Naturales", specially Mercedes Paris, ITACyL (Instituto Tecnológico Agrario de Castilla y León), Canterbury Museum, the New Zealand Arthropod Collection and the Museum of New Zealand Te Papa Tongarewa for access to historical collections. IB was funded by Fundación Banco Bilbao Vizcalla Argentaria (FBBVA) project. DW was funded through Landcare Research within the Characterising New Zealand’s Land Biota Portfolio.

>>>>

---
title: Historic collections as a tool for assessing the global pollination crisis
author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
csl: data/philb_style.xml
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
bibliography: data/references.xml
---

@@

For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1). Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).

@@

**Figure 1:** Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size. This correction accounts for the species-area relationship. Warmer colours indicate higher bee diversity. Note that some African countries may have incomplete listed faunas and that Alaska is included with US values. Countries with available historical changes in (b) bee, (c) syrphid and (d) butterfly richness within the last 100 years. Warmer colours indicate steeper average declines. Countries without data are coloured in white.

@@

#average richness change
#colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- (heat.colors(10))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

**Figure 2:** Exploration of available data for bee records showing (a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country. The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red. For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is slightly lower (red lines) in recent time periods (b). A more careful analysis of this data would help complete the map of global declines (c). In the map we plot % change is species recorded in gbif for the available countries to show the geographic coverage. Note that this data are likely to contain strong undetected biases, as we explore below.

@@

```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
dat <- read.csv("data/spain_clean_used.csv")
#Repeat richness and genus
rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#rich_tot #wow.
#extract genus
pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
dat$genus <- substr(dat$matched_name2, 
                    start = 1,
                    stop = pos-1)
genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot #wow, lot's of genus lost!
#genus_tot2
#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
              fun.aggregate = function(x) length(unique(x)))
#genus_tot3
#colSums(genus_tot3[,2:3])
#write.csv(genus_tot3, "data/SupMat3.csv")

@@

**Figure 3:** Comparision of historic collections (1980's) and modern re-surveys (2016) of the rarefied richness of bees at six Spanish localities.

@@

**Figure 4:** Changes in rarefied species richness for different pollinator groups in New Zealand over time. All trends were non-significant (α = 0.05).

@@

**Figure 5** Model estimated changes (± 1 SE) in the relative occurrence frequency of different New Zealand bee and fly species in museum collections over time.

@@

We thank Curro Molina, Carola Warner, Patrick McQuinn, and Crona McMonagle for data entry and Gregorio Aguado for carrying out the Spanish re-sampling. We thank Barry Donovan for providing New Zealand bee collection records and E. Asensio for sharing his historical data and knowledge. We thank the "Museo Nacional de Ciencias Naturales", specially Mercedes Paris, ITACyL (Instituto Tecnológico Agrario de Castilla y León), Canterbury Museum, the New Zealand Arthropod Collection and the Museum of New Zealand Te Papa Tongarewa for access to historical collections. IB was funded by Fundación Banco Bilbao Vizcaya Argentaria (FBBVA) project. DW was funded through Landcare Research within the Characterising New Zealand’s Land Biota Portfolio.
